`r glue::glue('Initial inferCNV results for {params$library_id}')`

### 2025-04-15 analyses/infercnv-consensus-cell-type/template-notebooks/ewings-infercnv-results.Rmd
updates in general, and specifically in the cnv section at the end

---
title: "`r glue::glue('Initial inferCNV results for {params$library_id}')`"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
params:
  sample_id: SCPCS000490
  library_id: SCPCL000822
---

@@

library(SingleCellExperiment)
library(ggplot2)
library(patchwork)

@@

```{r}
# First, construct data frame of total `has_cnv` counts, with other variables needed for plots
has_cnv_df <- infercnv_df |>
  dplyr::select(
    barcodes,
    normal_reference,
    subcluster,
    starts_with("has_cnv_")
  ) |>
  tidyr::pivot_longer(
    starts_with("has_cnv_"),
    names_to = "chr",
    values_to = "cnv"
  ) |>
  # sum all values for each cell, but separately for each run of inferCNV (aka each normal reference)
  dplyr::group_by(barcodes, normal_reference) |>
  dplyr::summarize(total_cnv_per_cell = sum(cnv)) |>
  # add indicator columns for whether the cell is reference or query
  dplyr::mutate(
    immune_reference_cell = barcodes %in% immune_ref_barcodes,
    subset_reference_cell = barcodes %in% subset_ref_barcodes
  ) |>
  # join in other data
  dplyr::left_join(celltype_df) |>
  dplyr::left_join(umap_df)
```

@@

```{r}
query_has_cnv_df |>
  # arbitrary filtering, this is just data
  dplyr::filter(normal_reference == "all-immune") |>
  ggplot() +
  aes(x = UMAP1, y = UMAP2, color = forcats::fct_lump(ewing_annotation, 8)) +
  geom_point(size = 0.1, alpha = 0.3) +
  facet_wrap(vars(normal_reference)) +
  labs(color = "annotation") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  umap_theme
```
Next, the total CNVs per cell; ideally, higher values will overlap with regions in the UMAP containing putative tumor cells.

```{r fig.width=10, warning = FALSE}
# add clusters into data frame
has_cnv_df <- infercnv_df |>
  dplyr::select(
    barcodes,
    normal_reference,
    infercnv_cluster = subcluster
  ) |>
  dplyr::inner_join(
    has_cnv_df,
    by = c("barcodes", "normal_reference")
  )

# noting we expect warnings with this plot since there's no UMAP coordinates
# for cells from other libraries in the reference
ggplot(has_cnv_df) +
  aes(x = UMAP1, y = UMAP2, color = total_cnv_per_cell) +
  geom_point(size = 0.1, alpha = 0.3) +
  scale_color_viridis_c() +
  facet_wrap(vars(normal_reference)) +
  labs(title = "Total CNV per cell") +
  umap_theme
```

@@

```{r}
compare_refs <- has_cnv_df |>
  # bring in library_id so we can only include those cells.
  dplyr::inner_join(
    infercnv_df |>
      dplyr::select(barcodes, library_id) |>
      unique()
  ) |>
  dplyr::filter(library_id == params$library_id) |>
  dplyr::select(
    barcodes, normal_reference, total_cnv_per_cell
  ) |>
  tidyr::pivot_wider(
    names_from = normal_reference,
    values_from = total_cnv_per_cell
  ) |>
  # color points this way rather than binning to ensure small bins aren't left out
  dplyr::add_count(`all-immune`, `subset-immune`)

@@

```{r}
ggplot(has_cnv_df) +
  # keep in default order so "reference" and "unknown" are grouped
  aes(x = infercnv_cluster) +
  geom_bar() +
  facet_wrap(
    vars(normal_reference),
    scales = "free_x"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

```{r fig.width=12}
ggplot(query_has_cnv_df) +
  aes(x = infercnv_cluster, fill = forcats::fct_lump(consensus_annotation, 7)) +
  geom_bar() +
  scale_fill_brewer(palette = "Dark2") +
  facet_wrap(
    vars(normal_reference),
    scales = "free_x"
  ) +
  labs(
    title = "Consensus cell types",
    fill = "annotation"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r fig.width=12}
ggplot(query_has_cnv_df) +
  aes(x = infercnv_cluster, fill = forcats::fct_lump(ewing_annotation, 7)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(
    vars(normal_reference),
    scales = "free_x"
  ) +
  labs(
    title = "Ewing cell types",
    fill = "annotation"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

```{r fig.width=10}
all_reference_clusters <- has_cnv_df |>
  # bring in library_id
  dplyr::inner_join(
    infercnv_df |>
      dplyr::select(barcodes, library_id) |>
      unique()
  ) |>
  dplyr::filter(
    normal_reference == "all-immune",
    immune_reference_cell
  )
all_plot <- ggplot(all_reference_clusters) +
  aes(x = infercnv_cluster, fill = library_id) +
  geom_bar() +
  labs("All immune reference") +
  theme(legend.position = "bottom")

subset_reference_clusters <- has_cnv_df |>
  # bring in library_id
  dplyr::inner_join(
    infercnv_df |>
      dplyr::select(barcodes, library_id) |>
      unique()
  ) |>
  dplyr::filter(
    normal_reference == "subset-immune",
    subset_reference_cell
  )
subset_plot <- ggplot(subset_reference_clusters) +
  aes(x = infercnv_cluster, fill = library_id) +
  geom_bar() +
  labs("Subset immune reference") +
  theme(legend.position = "bottom")

@@

```{r}
# How many CNV per cell per cluster?
ggplot(per_cell_cnv_df) +
  aes(x = infercnv_cluster, y = total_cnv_per_cell) +
  geom_point(size = 2) +
  facet_wrap(
    vars(normal_reference),
    scales = "free_x"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

Now, we'll look at CNVs across chromosomes, including specific CNVs we expect to see in Ewing sarcoma:

@@

```{r}
cnv_validation_df <- infercnv_df |>
  dplyr::select(
    barcodes,
    infercnv_cluster,
    total_cnv_per_cell,
    normal_reference,
    is_immune_reference,
    is_subset_reference,
    all_of(
      c(
        glue::glue("proportion_scaled_cnv_chr{1:22}"),
        glue::glue("proportion_scaled_loss_chr{1:22}"),
        glue::glue("proportion_scaled_dupli_chr{1:22}")
      )
    )
  ) |>
  tidyr::pivot_longer(
    starts_with("proportion_scaled_"),
    names_to = "cnv_type_raw",
    values_to = "proportion"
  ) |>
  dplyr::mutate(cnv_type_raw = stringr::str_remove(cnv_type_raw, "^proportion_scaled_")) |>
  tidyr::separate(cnv_type_raw, into = c("cnv_type", "chr")) |>
  dplyr::mutate(chr = stringr::str_remove(chr, "^chr"), chr = as.numeric(chr))
```

```{r}
cnv_validation_df |>
  dplyr::filter(normal_reference == "subset-immune") |>
  ggplot() +
  aes(
    x = proportion,
    color = is_subset_reference
  ) +
  geom_density() +
  facet_wrap(vars(chr))

cnv_validation_df |>
  dplyr::group_by(
    normal_reference,
    cnv_type,
    chr
  ) |>
  dplyr::summarize(median_proportion = median(proportion)) |>
  dplyr::arrange(cnv_type)
```

>>>>

---
title: "`r glue::glue('Initial inferCNV results for {params$library_id}')`"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
params:
  sample_id: SCPCS000490
  library_id: SCPCL000822
---

@@

suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(ggplot2)
  library(patchwork)
})

@@

```{r}
# First, construct data frame of total `has_cnv` counts, with other variables needed for plots
has_cnv_df <- infercnv_df |>
  dplyr::select(
    barcodes,
    normal_reference,
    starts_with("has_cnv_")
  ) |>
  tidyr::pivot_longer(
    starts_with("has_cnv_"),
    names_to = "chr",
    values_to = "cnv"
  ) |>
  # sum all values for each cell, but separately for each run of inferCNV (aka each normal reference)
  dplyr::group_by(barcodes, normal_reference) |>
  dplyr::summarize(total_cnv_per_cell = sum(cnv)) |>
  dplyr::ungroup() |>
  # add indicator columns for whether the cell is reference or query
  dplyr::mutate(
    immune_reference_cell = barcodes %in% immune_ref_barcodes,
    subset_reference_cell = barcodes %in% subset_ref_barcodes
  ) |>
  # join in other data
  dplyr::left_join(celltype_df) |>
  dplyr::left_join(umap_df)
```

@@

```{r, fig.width = 6}
query_has_cnv_df |>
  # arbitrary filtering, this is just data
  dplyr::filter(normal_reference == "all-immune") |>
  ggplot() +
  aes(x = UMAP1, y = UMAP2, color = forcats::fct_lump(ewing_annotation, 8)) +
  geom_point(size = 0.1, alpha = 0.3) +
  facet_wrap(vars(normal_reference)) +
  labs(color = "annotation") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  umap_theme
```

Next, the total CNVs per cell; ideally, higher values will overlap with regions in the UMAP containing putative tumor cells.

```{r fig.width=10, warning = FALSE}
# noting we expect warnings with this plot since there's no UMAP coordinates
# for cells from other libraries in the reference
ggplot(has_cnv_df) +
  aes(x = UMAP1, y = UMAP2, color = total_cnv_per_cell) +
  geom_point(size = 0.1, alpha = 0.3) +
  scale_color_viridis_c() +
  facet_wrap(vars(normal_reference)) +
  labs(title = "Total CNV per cell") +
  umap_theme
```

@@

```{r, fig.height=6}
compare_refs <- has_cnv_df |>
  # bring in library_id so we can only include those cells.
  dplyr::inner_join(
    infercnv_df |>
      dplyr::select(barcodes, library_id) |>
      unique()
  ) |>
  dplyr::filter(library_id == params$library_id) |>
  dplyr::select(
    barcodes, normal_reference, total_cnv_per_cell
  ) |>
  tidyr::pivot_wider(
    names_from = normal_reference,
    values_from = total_cnv_per_cell
  ) |>
  # color points this way rather than binning to ensure small bins aren't left out
  dplyr::add_count(`all-immune`, `subset-immune`)

@@

```{r, fig.width = 10}
# bring in cluster and library information
cluster_cnv_df <- has_cnv_df |>
  dplyr::inner_join(
    infercnv_df |>
      dplyr::select(barcodes, library_id, subcluster, normal_reference) |>
      unique(),
    by = c("barcodes", "normal_reference")
  )

ggplot(cluster_cnv_df) +
  # keep in default order so "reference" and "unknown" are grouped
  aes(x = subcluster) +
  geom_bar() +
  facet_wrap(
    vars(normal_reference),
    scales = "free_x"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

```{r fig.width=12}
cluster_cnv_df |>
  # remove cells from other libraries
  dplyr::filter(library_id == params$library_id) |>
  ggplot() +
  aes(x = subcluster, fill = forcats::fct_lump(consensus_annotation, 7)) +
  geom_bar() +
  scale_fill_brewer(palette = "Dark2") +
  facet_wrap(
    vars(normal_reference),
    scales = "free_x"
  ) +
  labs(
    title = "Consensus cell types",
    fill = "annotation"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r fig.width=12}
cluster_cnv_df |>
  # remove cells from other libraries
  dplyr::filter(library_id == params$library_id) |>
  ggplot() +
  aes(x = subcluster, fill = forcats::fct_lump(ewing_annotation, 7)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(
    vars(normal_reference),
    scales = "free_x"
  ) +
  labs(
    title = "Ewing cell types",
    fill = "annotation"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

```{r fig.width=10}
all_reference_clusters <- cluster_cnv_df |>
  dplyr::filter(
    normal_reference == "all-immune",
    immune_reference_cell
  )
all_plot <- ggplot(all_reference_clusters) +
  aes(x = subcluster, fill = library_id) +
  geom_bar() +
  labs("All immune reference") +
  theme(legend.position = "bottom")

subset_reference_clusters <- cluster_cnv_df |>
  # bring in library_id
  dplyr::inner_join(
    infercnv_df |>
      dplyr::select(barcodes, library_id) |>
      unique()
  ) |>
  dplyr::filter(
    normal_reference == "subset-immune",
    subset_reference_cell
  )
subset_plot <- ggplot(subset_reference_clusters) +
  aes(x = subcluster, fill = library_id) +
  geom_bar() +
  labs("Subset immune reference") +
  theme(legend.position = "bottom")

@@

```{r}
# How many CNV per cell per cluster?
ggplot(cluster_cnv_df) +
  aes(x = subcluster, y = total_cnv_per_cell) +
  geom_point(size = 2) +
  facet_wrap(
    vars(normal_reference),
    scales = "free_x"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

Now, we'll look at CNVs across chromosomes, specifically considering the `proportion_scaled_cnv_` output from the HMM.

Again, these are the specific CNVs we expect to see in Ewing sarcoma:

@@

```{r}
# First, construct data frame with proportion_scaled_ instead of has_cnv
cnv_validation_df <- infercnv_df |>
  dplyr::select(
    barcodes,
    library_id,
    normal_reference,
    starts_with("proportion_scaled_")
  ) |>
  # tidy up
  tidyr::pivot_longer(
    starts_with("proportion_scaled_"),
    names_to = "cnv_type_raw",
    values_to = "proportion"
  ) |>
  # more tidying
  dplyr::mutate(cnv_type_raw = stringr::str_remove(cnv_type_raw, "^proportion_scaled_")) |>
  tidyr::separate(cnv_type_raw, sep = "_", into = c("cnv_type", "chr")) |>
  # remove extra chromosomes and cells not in this library, and then can make chr numeric
  dplyr::filter(
    !stringr::str_starts(chr, "chrMT"),
    !stringr::str_starts(chr, "chrGL"),
    library_id == params$library_id
  ) |>
  dplyr::mutate(
    chr = stringr::str_remove(chr, "^chr"),
    chr = as.numeric(chr)
  )
```

We'll make density plots of the predicted CNV events across chromosomes, for each of the two normal references:

```{r fig.height=8, fig.width=8}
cnv_validation_df |>
  dplyr::filter(normal_reference == "all-immune") |>
  ggplot() +
  aes(
    x = proportion,
    color = cnv_type
  ) +
  geom_density() +
  facet_wrap(vars(chr)) +
  ggtitle("All immune reference")

cnv_validation_df |>
  dplyr::filter(normal_reference == "subset-immune") |>
  ggplot() +
  aes(
    x = proportion,
    color = cnv_type
  ) +
  geom_density() +
  facet_wrap(vars(chr)) +
  ggtitle("Subset immune reference")
```
